---
title: "Skew Symmetric Mixture"
author: " "
date: " "
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("dependencies.R")
library(foldcluster) # our R package
library(ggplot2)
library(mcclust)
library(mcclust.ext)
library(mclust)
library(grid)
library(gridExtra)
library(Rcpp)
library(RcppArmadillo)
library(foreach)
library(doParallel)
library(plyr)
library(dplyr)
library(reshape2)
```
# Introduction
This is the simulation study for the skew-symmetric mixture. In this example, we repeatedly sample from this mixture, then cluster the data using FOLD, VI, Binder's, and \texttt{mclust}. This implementation is done using the \texttt{foreach} package in order to speed up computation. If you just want to see the results of one replication from this method, see the file "nongauss_cluster.R" in the "extra_R" folder.

# Preliminaries
```{r}
# r packages
source("extra_R/nongaussmix.R")
source("extra_R/nongauss_cluster.R")
# parallel computing
n.cores <- 10
## create the cluster
my.cluster <- parallel::makeCluster(
  n.cores, 
  type = "PSOCK"
  )
## register it to be used by %dopar%
doParallel::registerDoParallel(cl = my.cluster)
```

# Simulation Study
```{r, warning = F}
N <- c(100, 500, 1000, 2500) # sample size
d <- 2 # dimension
R <- 100  # replications at each n
K_0 <- 3 # number of true mixture components
S <- 9000 # iterations
B <- 1000 # burnin
L <- 30 # number of fitted components
x <- foreach(i = 1:length(N)) %:%
  foreach(r = 1:R,
          .packages = c("mcclust", "mclust", "foldcluster", "mvtnorm", "sn")) %dopar% {
  nongauss.cluster(n = N[i],
               S = S,
               B = B,
               d = d,
               L = L)
                           }
Kplots <- ariplots <- Klist <- arilist <- list()
K_avg <- K_sd <- ari_avg <- ari_sd <- matrix(0, nrow = length(N), ncol = 4)
for (i in 1:length(N)){
  x_i <- x[[i]]
  K_i <- data.frame(matrix(unlist(x_i)[grepl("K", names(unlist(x_i)))], byrow = T, ncol = 4))
  Klist[[i]] <- K_i
  ari_i <- data.frame(matrix(unlist(x_i)[grepl("ari", names(unlist(x_i)))], byrow = T, ncol = 4))
  arilist[[i]] <- ari_i
  colnames(K_i) <- colnames(ari_i) <- c("FOLD", "VI", "Binder", "Mclust")
  # avgs
  K_avg[i,] <- round(colMeans(K_i),3)
  K_sd[i,] <- round(apply(K_i, 2, sd),3)
  ari_avg[i,] <- round(colMeans(ari_i),3)
  ari_sd[i,] <- round(apply(ari_i, 2, sd),3)
  # K plots
  m_Ki <- melt(K_i)
  Kplots[[i]] <- m_Ki %>% ggplot(aes(x = as.factor(variable), y = value, fill = as.factor(variable))) +   geom_boxplot() + xlab("Method") + ylab("No. of Clusters") +
    labs(title = paste("n=", toString(N[i]))) + 
    theme_bw() + 
    theme(legend.position = "none",
        text = element_text(size = 15)) 
  # ari plots
  m_ari <- melt(ari_i)
  ariplots[[i]] <- m_ari %>% ggplot(aes(x = as.factor(variable), y = value, fill = as.factor(variable)))  + geom_boxplot() + xlab("Method") + ylab("Adjusted Rand Index") +
    labs(title = paste("n=", toString(N[i]))) + 
    theme_bw() + 
    theme(legend.position = "none",
        text = element_text(size = 15)) 
}
grid.arrange(Kplots[[1]],
             Kplots[[2]],
             Kplots[[3]],
             Kplots[[4]],
             nrow = 2)
grid.arrange(ariplots[[1]],
             ariplots[[2]],
             ariplots[[3]],
             ariplots[[4]],
             nrow = 2)
# evaluating number of clusters
Kmat <- rbind(Klist[[1]],Klist[[2]],Klist[[3]],Klist[[4]])
m <- mean(Kmat[,1]<=Kmat[,2])
arimat <- rbind(arilist[[1]],arilist[[2]],arilist[[3]],arilist[[4]])
m_a <- mean(arimat[,1] >= arimat[,2])
```

